### 2017-02-14 thesis/methods/methods.tex
tmp commit
finished pathway and obesity metagene section - tmp commit

@@

In this project, two types of metagenes were created: obesity associated and pathway associated metagenes.
For obesity associated metagenes, the direction of the metagenes were checked with the obesity status and \gls{bmi} value of the samples from the data in which the metagene was derived from.
Checking the direction of the pathway metagenes from \citet{Gatza2010a} study is not as simple as the obesity associated metagenes, as the experimental design that identify which samples were treated to generate the pathway signature was not available.
Due to this, it was not possible to validate whether the generated pathway metagene was in the right direction.

As an alternative, the pathway metagenes were visually compared with the gene that represent the pathway using a heatmat, as the expression of this gene would most likely be affected the most by the pathway signature.
For example, the Akt pathway metagene was compared with the AKT1 gene expression, and if the metagene was not in the same direction as the AKT1 gene expression, the metagene was flipped.
\cref{tab:metagene_direction} summarises the genes used to check the direction of the pathway metagenes.
\begin{table}[htb]
	\caption{18 pathways from \citet{Gatza2010a} and its respective genes used to check the direction of the pathway metagene}
	\label{tab:metagene_direction}
	\begin{center}
		\begin{tabular}{lc}
			Pathway & Representative gene\\
			\hline
			\rule{0pt}{2.25ex}Akt & AKT1\\
			\gls{bcat}            & CTNNB1\\
			E2F1                  & E2F1\\
			\gls{egfr}            & EGFR\\
			\gls{er}              & ESR1\\
			\gls{her2}            & ERBB2\\
			\gls{ifna}            & IFNA1\\
			\gls{ifny}            & IFNG\\
			Myc                   & MYC\\
			p53                   & TP53\\
			p63                   & TP63\\
			\gls{pi3k}            & PIK3CA\\
			\gls{pr}              & PGR\\
			Ras                   & HRAS\\
			\gls{stat3}           & STAT3\\
			Src                   & SRC\\
			\gls{tgfb}            & TGFB1\\
			\gls{tnfa}            & TNF\\
			\hline
			\hline
		\end{tabular}
	\end{center}
\end{table}
In addition to this, the correlation of the pathway metagene with its repective gene was checked to further aid the decision on whether the metagene is going in the correct direction.
This was due to the fact that some of the metagenes were visually difficult to determine whether it was in the right direction.
Therefore, the direction as well as the magnitude of the correlation value was also taken into account in making the decision whether to flip the metagene.
Lastly, further on top of this, the clustering of the pathway metagenes with the other pathway metagenes were also considered.
It was known from \citet{Gatza2010a} study that some pathways clustered and grouped together with specific pathways, so the metagenes were flipped so that the grouping of the pathways matched the results from their study.

@@

>>>>

% TODO: add example plots for metagene direction

@@

In this project, two types of metagenes were created: obesity associated and pathway associated metagenes.
For obesity associated metagenes, the direction of the metagenes were checked with the obesity status and \gls{bmi} value of the samples from the data in which the metagene was derived from.
When the data set did not have \gls{bmi}/\gls{bmi} status information for the samples, a set of common genes were taken from all of the obesity associated genetic signatures and their expressions compared with the metagene scores.
Since all of the genetic signatures are associated with obesity, the genes that are common to all of the obesity associated genetic signatures will be related to the metagenes produced by these signatures, and therefore the expressions of these genes should be in line with the metagene scores.

Checking the direction of the pathway metagenes from \citet{Gatza2010a} study is not as simple as the obesity associated metagenes, as the experimental design that identify which samples were treated to generate the pathway signature was not available.
Due to this, it was not possible to validate whether the generated pathway metagene was in the right direction.
As an alternative, the pathway metagenes were visually compared with the gene that represent the pathway using a heatmat, as the expression of this gene would most likely be affected the most by the pathway signature.
For example, the Akt pathway metagene was compared with the AKT1 gene expression, and if the metagene was not in the same direction as the AKT1 gene expression, the metagene was flipped.
\cref{tab:metagene_direction} summarises the genes used to check the direction of the pathway metagenes.

\begin{table}[htb]
	\caption{18 pathways from \citet{Gatza2010a} and its respective genes used to check the direction of the pathway metagene}
	\label{tab:metagene_direction}
	\begin{center}
		\begin{tabular}{lc}
			Pathway & Representative gene\\
			\hline
			\rule{0pt}{2.25ex}Akt & AKT1\\
			\gls{bcat}            & CTNNB1\\
			E2F1                  & E2F1\\
			\Gls{egfr}            & EGFR\\
			\Gls{er}              & ESR1\\
			\Gls{her2}            & ERBB2\\
			\Gls{ifna}            & IFNA1\\
			\Gls{ifny}            & IFNG\\
			Myc                   & MYC\\
			p53                   & TP53\\
			p63                   & TP63\\
			\Gls{pi3k}            & PIK3CA\\
			\Gls{pr}              & PGR\\
			Ras                   & HRAS\\
			\Gls{stat3}           & STAT3\\
			Src                   & SRC\\
			\Gls{tgfb}            & TGFB1\\
			\Gls{tnfa}            & TNF\\
			\hline
			\hline
		\end{tabular}
	\end{center}
\end{table}

In addition to this, the correlation of the pathway metagene with its repective gene was checked to further aid the decision on whether the metagene is going in the correct direction.
This was due to the fact that some of the metagenes were visually difficult to determine whether it was in the right direction.
Therefore, the direction as well as the magnitude of the correlation value was also taken into account in making the decision whether to flip the metagene.
Lastly, the clustering of the pathway metagenes with the other pathway metagenes were also considered.
It was known from \citet{Gatza2010a} study that some pathways clustered and grouped together with specific pathways, so the metagenes were flipped so that the grouping of the pathways matched the results from their study.

@@

% TODO: talk about spearman correlation?

### 2017-02-14 thesis/results/results2.tex
finished pathway and obesity metagene section - tmp commit

% TODO: start from here
In contrast to the pathway associated genetic signatures, all of the obesity associated genetic signatures showed high correlation of the metagenes across the different data sets.

@@

>>>>

In contrast to the pathway associated genetic signatures, all of the obesity associated genetic signatures showed high correlation of the \gls{svd}- and transformation matrix-generated metagenes across the different data sets.
This suggested that obesity associated genetic signatures were consistent across all data sets and the transformation matrices for these signatures could be made in any of the data sets.
Taken together, transformation matrices for all the genetic  signatures were decided to be made in the Gatza's data set, since there were some pathway associated signatures that were specific to Gatza's data, but none of the obesity associated genetic signatures were specific to a data set.
\\

@@

\noindent
To visually determine which pathway associated genetic signatures were most similar to the obesity associated genetic signatures, heatmaps were created with the metagenes for all of the genetic signatures.
The directions of the pathway associated metagenes have already been determined in \cref{sec:pathway_associated_genetic_signatures_from_gatza2010a_study}, and the directions of the obesity associated metagenes were checked in Gatza's data set as described in \cref{sub:metagene_direction}.
All of the metagenes were created in \gls{rma}-normalised Gatza's data set with \gls{svd}, and the metagene scores were plotted in a heatmap and clustered into groups (\cref{fig:gatza_allmeta}).

\begin{figure}[htpb]
	\centering
	\includegraphics[page=15,width=0.8\linewidth]{results2/all_meta_trans}
	\caption{All metagenes in Gatza's data}
	\label{fig:gatza_allmeta}
\end{figure}

From \cref{fig:gatza_allmeta}, all of the obesity associated metagenes from Creighton's data set clustered together into a group, but none of the pathway associated metagenes associated strongly with the obesity metagenes.
Though \gls{her2} and p63 pathway metagenes were grouped next to the cluster of obesity metagenes, the similarity was at most about 0.6.
Fuentes-Mattei's obesity associated metagene did not cluster together with the obesity metagenes from Creighton's data set, nor with any of the other pathway metagenes.
The fact that Fuentes-Mattei's metagene did not cluster with Creighton's metagenes showed that the nature of Fuentes-Mattei's metagene was different from those metagenes found in Creighton's data set.
Furthermore, their metagene did not cluster with Gatza's Akt pathway in the heatmap, even though \citet{Fuentes-Mattei2014} provided clear evidence that their metagene specifically activated the Akt/\gls{mtor} signalling pathway in a mouse model.
This may have been due to the lack of consistency of the Akt pathway signature in different data sets (\cref{tab:svd_vs_tm_path}).
Another likely reason could be that Fuentes-Mattei's genetic signature was not comprised only of the genes from the Akt pathway, but genes from other pathways as well, and so the metagene did not cluster together with the Akt pathway signature in the heatmap.

The results in this section suggested that none of the obesity associated genetic signatures were similar or related to any of the pathway associated genetic signatures from the \citet{Gatza2010a} study.

(add heatmaps of other data as well? or maybe in appendix)

% Transformation matrices for all of the obesity and pathway associated genetic signatures were generated in Gatza's data.
